Dynamic Witnesses for Static Type Errors

### 2015-11-20 paper/pldi16-submission/interactive.tex
tweak intro
up to related

\section{Explaining Type Errors With Traces}
\label{sec:explaining}

@@

Next, we show how to build an visual interactive debugger
from the traced semantics, by describing the visualization
\emph{state} \ie\ what the user sees at any given moment,
the set of \emph{commands} available to user, and what
they do, and finally how we use a command to \emph{update}
the visualization state. In the sequel, for clarity of
exposition, we assume we have a (global) trace:
$\stepg{e_0}{\emptysu}{\bullet}{e_n}{\_}{\tr}$, where
$e_0$ and $e_n$ are the \emph{initial} and \emph{final}
expressions respectively.

\paragraph{Visualization State}
%
A \emph{visualization state} @VState@ is a \emph{directed graph}
whose vertices are expressions and whose edges are such
that each vertex has at most one predecessor and at most one
successor. In other words, the visualization state looks
like a set of linear lists of expressions as shown in Figure~\ref{fig:FIXTHIS}.
%
The \emph{initial state} is the graph containing a single
edge linking the initial and final expressions.

@@

\paragraph{Update}
Figure~\ref{fig:traversing-graph} shows how we update the state
for each command.
%
The procedure @findExpr vs cmd e@ traverses $\tr$
and the current visualization state to compute the new
expression that should be added to the visualization state
and @updState vs cmd e@ then updates the graph
by inserting the new expression appropriately, using one of
the following functions.
%
@putBefore vs e e'@ \hbox{(resp. @putAfter vs e e'@)}
returns the modified version of @vs@ where @e'@ is the
immediate predecessor of @e@ (resp. the immediate successor of @e@);
%
@putRoot vs e e'@ returns the modified version of @vs@ extended
with a new root vertex @e@ and its successor @e'@, and,
%
% @getRoot vs e@ returns the vertex (expression) obtained
% by transitively following the predecessors of @e@ until a
% source vertex.
%
@getNext vs e@ returns the immediate successor of @e@ in $\tr$.
%
@getPrev vs e@ computes the path @p@ between @e@ and its immediate
predecessor in the current visualization, and then returns @e@'s immediate
predecessor along @p@.
%
@getSubterms vs e@ traverses the sub-term edges to decompose an
expression into a list of sub-expressions paired with their context.
%
@applyCtx vs e ctx@ applies @ctx@ to @e@, traversing the sub-term edges
in reverse to find the super-term of @e@.
%
\hbox{@findApp vs e@} builds on top of @getSubterms@ to find the first
application sub-term (if any), \ie the first sub-term that looks like
$\eapp{v_1}{v_2}$.
%
@findVal vs e@ traverses the single-step edges to find the final value
that @e@ reduces to.
%
Note that the sub-term edges $\searrow$ allow us to decompose an
expression into a sub-expression and the surrounding context, thus
enabling the \stepintosym\ and \stepoversym\ traversals.

\RJ{EXAMPLE of interaction from overview goes here}

>>>>

\section{Explaining Type Errors With Traces}
\label{sec:interactive}

@@

Next, we show how to build a visual interactive debugger
from the traced semantics, by describing the visualization
\emph{state} \ie\ what the user sees at any given moment,
the set of \emph{commands} available to user and what
they do, and finally how we use a command to \emph{update}
the visualization state. In the sequel, for clarity of
exposition, we assume we have a (global) trace:
$\stepg{e_0}{\emptysu}{\bullet}{e_n}{\_}{\tr}$, where
$e_0$ and $e_n$ are the \emph{initial} and \emph{final}
expressions respectively.

\paragraph{Visualization State}
%
A \emph{visualization state} @VState@ is a \emph{directed graph}
whose vertices are expressions and whose edges are such
that each vertex has at most one predecessor and at most one
successor. In other words, the visualization state looks
like a set of linear lists of expressions as shown in Figure~\ref{fig:nanomaly-factorial}.
%
The \emph{initial state} is the graph containing a single
edge linking the initial and final expressions.

@@

\paragraph{Update}
Figure~\ref{fig:traversing-graph} shows how we update the state
for each command.
%
The procedure @findExpr vs cmd e@ traverses $\tr$
and the current visualization state to compute the new
expression that should be added to the visualization state,
and @updState vs cmd e@ then updates the graph
by inserting the new expression appropriately, using one of
the following functions.
%
@putBefore vs e e'@ \hbox{(resp. @putAfter vs e e'@)}
returns the modified version of @vs@ where @e'@ is the
immediate predecessor of @e@ (resp. the immediate successor of @e@);
%
@putRoot vs e e'@ returns the modified version of @vs@ extended
with a new root vertex @e@ and its successor @e'@.
%
% @getRoot vs e@ returns the vertex (expression) obtained
% by transitively following the predecessors of @e@ until a
% source vertex.

@getNext vs e@ returns the immediate successor of @e@ in $\tr$.
%
@getPrev vs e@ computes the path @p@ between @e@ and its immediate
predecessor in the current visualization, and then returns @e@'s immediate
predecessor along @p@.
%
@getSubterms vs e@ traverses the sub-term edges to decompose an
expression into a list of sub-expressions paired with their context.
%
@applyCtx vs e ctx@ applies @ctx@ to @e@, traversing the sub-term edges
in reverse to find the super-term of @e@.
%
\hbox{@findApp vs e@} builds on top of @getSubterms@ to find the first
application sub-term (if any), \ie the first sub-term that looks like
$\eapp{v_1}{v_2}$.
%
@findVal vs e@ traverses the single-step edges to find the final value
that @e@ reduces to.
%
% Note that the sub-term edges $\searrow$ allow us to decompose an
% expression into a sub-expression and the surrounding context, thus
% enabling the \stepintosym\ and \stepoversym\ traversals.

% \RJ{EXAMPLE of interaction from overview goes here}
\begin{figure}[t]
\centering
\includegraphics[width=0.8\linewidth]{fac-steps.png}
\caption{A sequence of interactions with the trace of
  \texttt{fac 1}. The stuck term is red, in each node the redex is
  highlighted. Thick arrows denote a multi-step transition, thin arrows
  denote a single-step transition. We start in step 1. In step 2 we jump
  forward from the witness to the next function call. In step 3 we step
  into the recursive \texttt{fac 0} call, which spawns a new ``thread''
  of execution. In step 4 we take a single step forward from
  \texttt{fac 0}.} % (hiding the context for space).}
\label{fig:nanomaly-factorial}
\end{figure}

### 2015-11-20 paper/pldi16-submission/overview.tex
up to related

A sample interaction with the trace of @fac 1@ can be seen in
Figure~\ref{fig:nanomaly-factorial}.
%
\begin{figure*}[t]
\centering
\includegraphics[width=0.8\linewidth]{fac-steps.png}
\caption{A sequence of interactions with the trace of
  \texttt{fac 1}. The stuck term is red, in each node the redex is
  highlighted. Thick arrows denote a multi-step transition, thin arrows
  denote a single-step transition. We start in step 1. In step 2 we jump
  forward from the witness to the next function call. In step 3 we step
  into the recursive \texttt{fac 0} call, which spawns a new ``thread''
  of execution. In step 4 we take a single step forward from
  \texttt{fac 0} (hiding the context for space).}
\label{fig:nanomaly-factorial}
\end{figure*}
%
The initial state of the visualization tells us that after some number
of steps -- the thick arrow denotes a multi-step transition -- we try to
multiply @1@ by @true@.

Upon seeing the stuck term, we might wonder where @true@ came from.
%
To investigate we select the stuck term and click the ``jump backward''
button to search backwards from the stuck term for the most recent
function call, which brings us to @1 * fac 0@. Notice at this point that
@fac 0@ is highlighted while @1 *@ is grayed out. This tells us that
@fac 0@ is the redex in this term.

@fac 0@ seems like the right thing to do so we choose to ``step into''
it, which inserts a new multi-step transition from @fac 0@ to @true@.
%
Finally, we take a ``step forward'' from \hbox{@fac 0@,} bringing us to @fac@'s
body. If we mouse over the body term we will see a popup with the
environment at this point, notably telling us that @n = 0@. At this point
it is clear that @fac@ handles the @n <= 0@ case incorrectly and should
instead return an @int@.

>>>>

% A sample interaction with the trace of @fac 1@ can be seen in
% Figure~\ref{fig:nanomaly-factorial}.
% %
% % \begin{figure*}[t]
% % \centering
% % \includegraphics[width=0.8\linewidth]{fac-steps.png}
% % \caption{A sequence of interactions with the trace of
% %   \texttt{fac 1}. The stuck term is red, in each node the redex is
% %   highlighted. Thick arrows denote a multi-step transition, thin arrows
% %   denote a single-step transition. We start in step 1. In step 2 we jump
% %   forward from the witness to the next function call. In step 3 we step
% %   into the recursive \texttt{fac 0} call, which spawns a new ``thread''
% %   of execution. In step 4 we take a single step forward from
% %   \texttt{fac 0} (hiding the context for space).}
% % \label{fig:nanomaly-factorial}
% % \end{figure*}
% %
% The initial state of the visualization tells us that after some number
% of steps -- the thick arrow denotes a multi-step transition -- we try to
% multiply @1@ by @true@.

% Upon seeing the stuck term, we might wonder where @true@ came from.
% %
% To investigate we select the stuck term and click the ``jump backward''
% button to search backwards from the stuck term for the most recent
% function call, which brings us to @1 * fac 0@. Notice at this point that
% @fac 0@ is highlighted while @1 *@ is grayed out. This tells us that
% @fac 0@ is the redex in this term.

% @fac 0@ seems like the right thing to do so we choose to ``step into''
% it, which inserts a new multi-step transition from @fac 0@ to @true@.
% %
% Finally, we take a ``step forward'' from \hbox{@fac 0@,} bringing us to @fac@'s
% body. If we mouse over the body term we will see a popup with the
% environment at this point, notably telling us that @n = 0@. At this point
% it is clear that @fac@ handles the @n <= 0@ case incorrectly and should
% instead return an @int@.
